---
title: "Differential expression analysis"
author: "Scott Campit"
output: 
  html_notebook:
    toc: yes
---

## Summary
This notebook performs differential expression analysis for various transcriptomics datasets. 

**UPDATES: **

  * November 4, 2020: Added more stringent and robust differential expression analysis options.
  * Follow computational pipeline outlined in Yu, et al. (2019). Comprehensive transcriptomic analysis of cell lines as models of primary tumors across 22 tumor types. Nature communications, 10(1), 1-11.
  * Compute differentially expressed genes using both edgeR, DESeq2, and limma-voom
  * Take intersection of differentially expressed genes 
  
## 1. Load libraries
First we'll load some essential packages needed to run differential expression analysis.
```{r, warning=FALSE, message=FALSE}
# Bioinformatics libraries
#install.packages("BiocManager")
#BiocManager::install(c("BiocManager", "edgeR", 
#                   "DESeq2", "limma", 
#                   "sva", "PCAtools",
#                   "Glimma", "bioMart"))

#install.packages("CePa", dependences=TRUE)
#BiocManager::install("Rgraphviz")
library(CePa)
library(edgeR)
library(DESeq2)
library(limma)
library(sva)
library(PCAtools)
library(Glimma)
library(biomaRt)

# Data science libraries
#install.packages(c("tidyverse", "readr", "rrcov"))
library(tidyverse)
library(readr)
library(rrcov)

remove(list=ls())
```

## 2. Perform data preprocessing

This code block performs essential data preprocessing for the cancer cell line encyclopedia data.

### Read in raw RNASeq counts

First, let's load the table with the raw RNASeq gene counts.
```{r}
# ACLX path
#readpath = "C:\\Users\\scott\\Data\\RNASeq\\CCLE\\CCLE_934_TPM_EBI.tsv"
readpath = "C:\\Users\\scott\\Data\\RNASeq\\CCLE\\CCLE_RNAseq_genes_counts_20180929.gct"

# Dell
#readpath = "D:\\Data\\RNASeq\\CCLE\\CCLE_934_TPM_EBI.tsv"

# Load data either as a TPM or gene counts. I'm doing this from scratch.
#ccle = read_delim(file=readpath, delim='\t')

# Linux
#readpath = "~/Data/RNASeq/CCLE/CCLE_RNAseq_genes_counts_20180929.gct"
ccle = read.gct(readpath)
```

Note that in the raw gene count data matrix, the rows are labeled by Ensembl IDs, while the columns have the syntax CELLLINE_TISSUE, separated by an underscore. I can keep the Ensembl gene IDs, but I need to extract the tissue type, as this is a covariate of interest.

### Map genes to Entrez ID
First, since we'll be using Entrez ID in the RECON1 model, we need to map ENSEMBL to Entrez.
```{r}
# Get map of Ensembl to Entrez
human = useMart("ensembl", dataset="hsapiens_gene_ensembl")
genes = getBM(
  attributes=c("ensembl_gene_id", "hgnc_symbol", "entrezgene_id"),
  values=rownames(ccle),
  mart=human
)

# Map Entrez to CCLE dataset. Remove those without Entrez ID.
rownames(ccle) <- gsub("\\..*","",rownames(ccle))

# Map rows
mapped_ccle = merge(genes, ccle, by.x="ensembl_gene_id", by.y=0)

# Remove datapoints where entrez is NA
remove_genes = is.na(mapped_ccle[, "entrezgene_id"])
ccle = mapped_ccle[!remove_genes,]
ccle = ccle[, -c(1:2)]
agg_ccle = aggregate(ccle[, -1], 
                     by=list(ccle[, 1]), 
                     FUN=median)
colnames(agg_ccle)[1] = "entrezgene_id"
ccle = agg_ccle

rm(list = c("human", "genes", 
            "mapped_ccle", "agg_ccle", 
            "remove_genes"))
```

### Aggregate and remove duplicate cell line entries.
Note that some cell lines (columns) are duplicate entries. I need to aggregate them before moving on to additional preprocessing steps. 

Now let's begin preprocessing by grabbing some important meta data, specifically the tissue names.
```{r}
# Get data from the ccle dataframe
tmp = colnames(ccle)
tissue = word(string=tmp, start=2, end=-1, sep="_")
ccl = word(string=tmp, start=1, sep="_")

# Store into dataframe
colnames(ccle) = as.character(ccl)
```

Now we can being aggregating by cancer cell line.
```{r}
# Get entrez gene id
tmp = t(ccle)
entrezgene = tmp[1, ]
tmp = tmp[-1, ]

ccl = rownames(tmp)
rownames(tmp) = NULL

tissue = tissue[-1]
meta = data.frame("CCL"=ccl, "Tissue"=tissue)
meta = meta[!duplicated(ccl), ]

# Aggregate by cancer cell line
agg_tmp = aggregate(tmp, by=list(ccl), FUN=mean)
agg_tmp = t(agg_tmp)

agg_tmp = as.data.frame(agg_tmp)
colnames(agg_tmp) = agg_tmp[1, ]
agg_tmp = agg_tmp[-1, ]
rownames(agg_tmp) = ccle[, 1]
ccle = agg_tmp
rm(list = c("agg_tmp"))
```

### Missing value imputation
Note that there are missing values, so we'll impute them with the minimum value in the dataset.
```{r}
min_impute = function(df){
  for (cols in colnames(df)) {
    if (cols %in% names(df[,sapply(df, is.numeric)])) {
      df = df %>% mutate(!!cols := replace(!!rlang::sym(cols),
                                           is.na(!!rlang::sym(cols)),
                                           min(!!rlang::sym(cols), na.rm=TRUE)))
    }
  }
  return(df)
}

tmp = DGEList(counts=data.matrix(ccle), genes=rownames(ccle))
keep = rowSums(cpm(tmp) >= 1) >= (ncol(tmp$counts)/5)
ccle = ccle[keep, ]
ccle = min_impute(ccle)
```



### Separate solid tumor and hematopoietic cancer cell lines
I need to separate the hematopoeitic and leukemia cell line lineages. Then I can perform differential expression analysis on both of the them separately.
```{r}
HL_idx = (meta[, 2] == "HAEMATOPOIETIC_AND_LYMPHOID_TISSUE")
HL_CCL = ccle[ , c(HL_idx)]
HL_tissue = meta[c(HL_idx), 2]
SOLID_CCL = ccle[ , !c(HL_idx)]
SOLID_tissue = meta[!c(HL_idx), 2]
```

### Remove outliers
We will first remove some outliers in both the HL and SOLID datasets using Robust PCA. It combines two ideas to find outliers:
  1. Find projections with maximal dispersion using a grid search.
  2. Estimate the covariance using an idea similar to Hubert robust regression.

First, we'll define a function to do this filtering.
```{r}
# Create a function that filters data using ROBPCA
removeOutliers = function(data, id){
  pca_obj = PcaHubert(data)
  filtered_data = data[pca_obj@flag, ]
  filtered_id = id[pca_obj@flag]
  outlier_data = data[!pca_obj@flag, ]
  outlier_id = id[!pca_obj@flag]
  
  setClass(Class="geneObj",
           representation(
             data="matrix",
             outlier_data="matrix",
             id="character",
             flag="logical",
             outlier_id="character"
           )
   )
   return(new("geneObj",
          data=filtered_data,
          outlier_data=outlier_data,
          id=filtered_id,
          flag=pca_obj@flag,
          outlier_id=outlier_id))
}
```

Now let's do this filtering for both genes and cell lines.
```{r}
# Construct the input for PcaHubert of class matrix
HL_data = data.matrix(HL_CCL)
SOLID_data = data.matrix(SOLID_CCL)

# Use removeOutliers to remove noisy genes
HL_geneFilter = removeOutliers(HL_data, rownames(HL_data))
SOLID_geneFilter = removeOutliers(SOLID_data, rownames(SOLID_data))
```

```{r}
# Now set the stage to remove noisy cancer cell lines
HL_data2 = data.matrix(HL_geneFilter@data)
SOLID_data2 = data.matrix(SOLID_geneFilter@data)

HL_data2 = t(HL_data2)
SOLID_data2 = t(SOLID_data2)

HL_allFilter = removeOutliers(HL_data2, meta[, 1])
SOLID_allFilter = removeOutliers(SOLID_data2,  meta[, 1])
```

The final dataframe will be constructed here.
```{r}
# Construct the final dataframe
HL_final = t(data.matrix(HL_allFilter@data))
SOLID_final = t(data.matrix(SOLID_allFilter@data))
rownames(HL_final) = c(HL_geneFilter@id)
rownames(SOLID_final) = c(SOLID_geneFilter@id)
final_tissue = SOLID_tissue[SOLID_allFilter@flag]
```

If we want to do a final sanity check, here's a code block that can make those plots.
```{r}
tmp = t(data.matrix(SOLID_final))
colnames(tmp) = NULL
rownames(tmp) = NULL

# Visualize cancer cell lines in 2D space
solid_pc_filtered = PCAtools::pca(tmp)
PCAtools::screeplot(solid_pc_filtered)
PCAtools::biplot(solid_pc_filtered, 
                 x="PC1", y="PC2")
PCAtools::pairsplot(solid_pc_filtered)
```


```{r}
# Visualize genes in 2D space
tmp = data.matrix(SOLID_final)
colnames(tmp) = NULL
rownames(tmp) = NULL

solid_pc_filtered = PCAtools::pca(tmp)
PCAtools::screeplot(solid_pc_filtered)
PCAtools::biplot(solid_pc_filtered)
PCAtools::pairsplot(solid_pc_filtered)
```

### Batch Effect Correction using `ComBat-Seq`
Now we can use Combat-Seq to perform batch correction from our expression set. Specifically, we'll be correcting for tissue lineage relationships with the cleaned up dataset.

First, let's construct the `pheno` dataframe, a dataframe that contains

First let's construct the model matrix which contains the adjustment variables and the variable of interest.
```{r}
#mod = model.matrix(~as.factor(tissue), data=pheno)
#mod0 = model.matrix(~1, data=pheno)
```

Now let's compute the latent factors that need to be estimated. Note that this code will take awhile as it needs to perform several SVD operations.
```{r}
#n.sv = num.sv(dat=Xcpm, mod=mod)
#svobj = svaseq(Xcpm, mod, mod0, n.sv=n.sv)
```

We can actually get differentially expressed genes using a parametric F-test. Let's check this out!
```{r}
# Compute initial P and Q-values
#tmp_pval = f.pvalue(data.matrix(Xfinal), mod, mod0)
#tmp_qval = p.adjust(tmp_pval, method='BH')

# Create adjusted models
#modSv = cbind(mod, svobj$sv)
#mod0Sv = cbind(mod0, svobj$sv)

# Adjust p-values using Combat
#combat_pvalues = f.pvalue(data.matrix(Xfinal), modSv, mod0Sv)
#combat_qvalues = p.adjust(combat_pvalues, method='BH')
```

### Data Normalization 
Now let's take the log2 of the counts-per-million of the RNASeq counts.
```{r}
Xfinal = SOLID_final
Xfinal = DGEList(Xfinal, 
                 genes=rownames(Xfinal), 
                 group=as.factor(final_tissue))
Xfinal = calcNormFactors(Xfinal, method="TMM")
Xfinal = cpm(Xfinal, log=TRUE)
#Xfinal = Xfinal / rowMeans(Xfinal)
hist(Xfinal, breaks=100)
```

## 3. Differential Expression Analysis

### Log2 fold change
We'll use the log2 fold change 
```{r}
Xfinal = SOLID_final
mu = rowMeans(Xfinal)
sd = rowSds(Xfinal)
Xfinal = (Xfinal - mu) / sd
hist(Xfinal, breaks=100)
pvalue=2*pnorm(-abs(Xfinal))

tmp = as.vector(pvalue)
tmp = p.adjust(tmp, method="fdr")
pvalue = matrix(tmp, 
                nrow=nrow(pvalue), ncol=ncol(pvalue), 
                byrow=TRUE)
colnames(pvalue) = colnames(Xfinal)
rownames(pvalue) = rownames(Xfinal)

install.packages("openxlsx")
library(openxlsx)

df_list = list("Z-scores"=Xfinal, "P-value"=pvalue)
filename = "C:/Users/scott/Data/RNASeq/CCLE/DEG_RNASEQ_120920.xlsx"
write.xlsx(df_list, file=filename, row.names=TRUE)
```

### Perform *limma-voom* 
```{r}
design = model.matrix(~as.factor(final_tissue))
vt = voom(Xfinal, design, plot=TRUE)
vt_fit = lmFit(vt, design=design, method="robust")
vt_fit = eBayes(vt_fit, robust=TRUE, trend=TRUE)
limma_results = topTable(vt_fit, adjust.method="BH")
```

### Perform *EdgeR*
```{r}
disperse = estimateCommonDisp(Xnorm)
disperse = estimateTagwiseDisp(disperse)
dgeTest = exactTest(disperse, dispersion="auto", rejection.region="doubletail")
```

### Perform *DESeq2*
```{r}
deseqTable = DESeqDataSetFromMatrix(
  countData=data.matrix(SOLID_final),
  colData=colnames(SOLID_final),
  design=design
)
deseqTable = DESeq(deseqTable)
deseq2_results = results(deseqTable)
```