---
title: "Dimension reduction for CCLE dataset"
output: html_notebook
---

# Summary
This notebook performs dimension reduction on the CCLE global chromatin profiling data and the metabolomics data. I will perform both linear and non-linear dimension reduction.

# Library and data preprocessing
First we'll load some essential libraries.
```{r}
#install.packages(c('ggplot2', 'tidyverse', 'readxl'))
library(tidyverse)
library(ggplot2)
library(readxl)
library(reshape2)
```

Now let's load the data.
```{r}
filepath = 'C:/Users/scott/Data/Mapped CCLE Metabolomics-GCP Data.xlsx'
met = read_xlsx(path=filepath, sheet='Metabolomics')
celllines = met[, 1]
met = met[, -1]
met = data.matrix(met)

gcp = read_xlsx(path=filepath, sheet='GCP')
gcp = gcp[, -1]
gcp = data.matrix(gcp)

meta = read_xlsx(path=filepath, sheet='Meta')
```

## Clean dataset

I'm only interested in single histone PTMs, so I need to clean up the GCP dataset.
```{r}
gcp_name = colnames(gcp)
pos_to_keep = c(3, 4, 5, 
                7, 8, 9, 10, 
                11, 17, 18, 
                22, 23, 24, 
                25, 29, 32, 
                34, 42, 43) - 1
gcp_name = gcp_name[pos_to_keep]
```

Let's clean up the labels using regular expressions.
```{r}
# Remove useless marks
gcp = gcp[, pos_to_keep]

# Remove regexes
pattern = "K....0"
gcp_name = str_replace(gcp_name, pattern, '')

pattern = "K...0"
gcp_name = str_replace(gcp_name, pattern, '')

colnames(gcp) = gcp_name
```

We should sanity check this now with histograms
```{r}
ggplot(melt(met), aes(x=value)) + 
  geom_histogram(bins=100) + 
  ggtitle("CCLE Metabolomics distribution")

ggplot(melt(gcp), aes(x=value)) + 
  geom_histogram(bins=100) + 
  ggtitle("CCLE Global Chromatin Profile distribution")
```
Also, I should remove NAs from the data
```{r}
gcp[is.na(gcp)] = 0
```

Now, let's separate hematopoietic and leukemic stem cell lineages, just because they have different profiles.

```{r}
HL_idx = meta['Tissue'] == "HAEMATOPOIETIC AND LYMPHOID TISSUE"

# Solid only
gcp_noHL = gcp[-c(HL_idx), ]
met_noHL = met[-c(HL_idx), ]

# HL only
gcp_HL = gcp[c(HL_idx), ]
met_HL = met[c(HL_idx), ]

tissue = meta["Tissue"]
noHL_tissue = tissue[-c(HL_idx), 1]
HL_tissue = tissue[c(HL_idx), 1]
```

## Mean-center metabolomics dataset
The metabolomics dataset is not mean centered. I will do that now.
```{r}
met_noHL = met_noHL - mean(met_noHL)
met_HL = met_HL - mean(met_HL)
```

We should sanity check this now with histograms
```{r}
ggplot(melt(gcp_noHL), aes(x=value)) + 
  geom_histogram(bins=100) + 
  ggtitle("Solid tumor Global Chromatin Profile distribution")

ggplot(melt(met_noHL), aes(x=value)) + 
  geom_histogram(bins=100) + 
  ggtitle("Solid tumor Metabolomics distribution")

ggplot(melt(gcp_HL), aes(x=value)) + 
  geom_histogram(bins=100) + 
  ggtitle("HL Global Chromatin Profile distribution")

ggplot(melt(met_HL), aes(x=value)) + 
  geom_histogram(bins=100) + 
  ggtitle("HL Metabolomics distribution")
```

There are definitely still outliers in the GCP dataset, but We'll address those next.

## Outliers

Let's use robust PCA to identify cancer cell lines that are outliers.
```{r}
#install.packages('rrcov')
library(rrcov)

# Run Robust PCA
gcp_pca = PcaHubert(gcp)

# Remove outliers. We need to remove cell lines from the metabolomics dataset too.
gcp_noHL = gcp_noHL[gcp_pca@flag, ]
gcp_HL = gcp_HL[gcp_pca@flag, ]
met_noHL = met_noHL[gcp_pca@flag, ]
met_HL = met_HL[gcp_pca@flag, ]

celllines = celllines[gcp_pca@flag,]
```

Now let's plot the resulting data matrices.
```{r}
# Sanity Check
ggplot(melt(met), aes(x=value)) + 
  geom_histogram(bins=100) + 
  ggtitle("CCLE Metabolomics distribution") 

ggplot(melt(gcp), aes(x=value)) + 
  geom_histogram(bins=100) + 
  ggtitle("CCLE Global Chromatin Profile distribution")
```

# Correlation analysis with filtered GCP and MET matrices
Let's see how the data is correlated now that we removed bad data points.
```{r}
install.packages(c("corrplot", "RColorBrewer"))
library(corrplot)
library(RColorBrewer)
```

First, well compute the correlation matrix
```{r}
pcc = cor(met, gcp)
```

Now let's create a function that computes the p-value for the pearson correlation coefficient.
```{r}
cortest = function(df1, df2){
  pvalue = matrix(, nrow=ncol(df1), ncol=ncol(df2))
  for(i in 1:ncol(df1)){
    for(j in 1:ncol(df2)){
      pvalue[i, j] = cor.test(df1[, i], df2[, j])$p.value
    }
  }
  colnames(pvalue) = colnames(df2)
  rownames(pvalue) = colnames(df1)
  return(pvalue)
}
pmat = cortest(met, gcp)
```

Finally, we'll plot the correlation matrix.
```{r}
corrplot(pcc, 
         type="upper", 
         tl.col="black", tl.srt=45,
         tl.pos = "td", tl.cex = 0.5, 
         p.mat=pmat, sig.level=0.05)
```
# Conclusions
All of the data is saved on 
